Generating Correlated Data for Omics Simulation

### 2025-06-17 manuscript/paper.qmd
Edit MS and figures
And fix metabolomics figure

---
title: Generating Correlated Data for Omics Simulation
author:
    - name: "Jianing Yang"
      orcid: 0000-0002-2048-9398
      affiliations:
        - name: "Chronobiology and Sleep Institute, University of Pennsylvania"
    - name: "Gregory R. Grant"
      orcid: 0000-0002-0139-7658
      affiliations:
        - name: "Department of Genetics, University of Pennsylvania"
    - name: "Thomas G. Brooks"
      orcid: 0000-0002-6980-0079
      affiliations:
        - name: "Institute for Translational Medicine and Therapeutics, University of Pennsylvania"
editor: 
  markdown: 
    wrap: sentence
execute:
  echo: false
  cache: true
  warning: false
format:
  html:
    toc: true
    toc-expand: true
bibliography: references.bib
---
## Abstract

@@

We start with a reference dataset $X$ given by an $p \times n$ data matrix of $p$ features measured in each of $n$ independent samples, which could be any real dataset of interest.
Our goal is to generate simulated dataset that looks, at least on average, the same as this reference dataset.
To do this, we need to capture correlations between the $p$ features, which could represent gene expressions, protein abundances, or other measured values.
We refer to these features as genes for simplicity.

@@

Finally, we ran DESeq2 on each simulated 5 vs 5 experiment and compared the output FDR with the true percentages of genes that are differential expressed (@fig-DESeq2 a-d).
We observed that DESeq2 is anti-conservative on all datasets, with similar mean true FDRs for each estimated FDR cutoff.
However, we note that this anti-conservative nature is most present at very low FDR thresholds, was typically associated with only a single digit number of false positive genes.
While that mean behavior was consistent whether or not the simulation included gene-gene dependence, simulations with dependence showed a substantially greater variance in the performance of DESeq2.

@@

We demonstrated the importance of including gene-gene dependence in simulated data by two application benchmarks.
In the first, DESeq2 results were substantially more variable when simulating with gene-gene dependence.
This indicates that DESeq2 had an increased chance of a larger simultaneous number of false positives, as well as increased chance of fewer false positives, when data included realistic dependence.
We therefore recommend that benchmarking of differential expression methods should include gene-gene dependence whenever possible.
This is despite the fact that differential expression is largely addressing a problem of the behavior of individual genes.
In the second, CYCLOPS performance in estimating circadian phases depends upon gene-gene dependence, was sensitive to dependence structure of the data.
Unlike DESeq2, CYCLOPS explicitly makes use of the correlations of genes and therefore this result is not surprising but still demonstrates the potential impact of the assumption of independence when benchmarking.

@@

@@

While we describe one special form of $\Sigma_{sim}$ where generation is highly efficient, there exist many other possible covariance matrix forms.
Block diagonal covariance matrices are one such example which can efficiently represent even high-dimensional data [@Devijver02012018] and lend themselves to efficient simulation.
Sparse matrix methods [@Bien2011-nx, @Bickel_sparse, @Cai_adaptive] yield exactly zero correlation between some variables, but it is less clear how to simulate data from these efficiently.
Iterative sampling methods [@Krylov_subspaces] are one approach that could exploit the sparse nature and may even extend to sparse inverse covariance methods [@sparse_inverse_covariance].

>>>>

---
title: Generating Correlated Data for Omics Simulation
author:
    - name: "Jianing Yang"
      orcid: 0000-0002-2048-9398
      affiliations:
        - name: "Chronobiology and Sleep Institute, University of Pennsylvania"
    - name: "Gregory R. Grant"
      orcid: 0000-0002-0139-7658
      affiliations:
        - name: "Department of Genetics, University of Pennsylvania"
    - name: "Thomas G. Brooks"
      orcid: 0000-0002-6980-0079
      affiliations:
        - name: "Institute for Translational Medicine and Therapeutics, University of Pennsylvania"
editor: 
  markdown: 
    wrap: sentence
execute:
  echo: false
  cache: true
  warning: false
format:
  docx:
    fig-dpi: 800
  html:
    toc: true
    toc-expand: true
bibliography: references.bib
---
## Abstract

@@

We start with a reference dataset $X$ given by an $p \times n$ data matrix of $p$ features measured in each of $n$ independent samples, which could be any real dataset of interest.
Our goal is to generate a simulated dataset that looks, at least on average, the same as this reference dataset.
To do this, we need to capture correlations between the $p$ features, which could represent gene expressions, protein abundances, or other measured values.
We refer to these features as genes for simplicity.

@@

Finally, we ran DESeq2 on each simulated 5 vs 5 experiment and compared the output FDR with the true percentages of genes that are differential expressed (@fig-DESeq2 a-d).
We observed that DESeq2 is anti-conservative on all datasets, with similar mean true FDRs for each estimated FDR cutoff.
While the mean behavior was consistent whether or not the simulation included gene-gene dependence, simulations with dependence showed a substantially greater variance in the performance of DESeq2.

@@

We demonstrated the importance of including gene-gene dependence in simulated data by two application benchmarks.
In the first, DESeq2 results were substantially more variable when simulating with gene-gene dependence.
This indicates that DESeq2 had an increased chance of a larger simultaneous number of false positives, as well as increased chance of fewer false positives, when data included realistic dependence.
We therefore recommend that benchmarking of differential expression methods should include gene-gene dependence whenever possible.
This is despite the fact that differential expression is largely addressing a problem of the behavior of individual genes.
We note that DESeq2's observed anti-conservative nature is most present at very low FDR thresholds and was typically associated with only a single digit number of false positive genes.
Moreover, more modern methods to control FDR are compatible with DESeq2 and have been shown to be appropriately conservative [@Korthauer2019-lx].

In the second application, CYCLOPS performance in estimating circadian phases depends upon gene-gene dependence, was sensitive to dependence structure of the data.
Unlike DESeq2, CYCLOPS explicitly makes use of the correlations of genes and therefore this result is not surprising but still demonstrates the potential impact of the assumption of independence when benchmarking.

@@

We observe that our PCA method generally overshoots the level of correlation in the real data set.
The corpcor method typically undershoots.
The spiked Wishart method also undershoots but less so than the corpcor method and therefore we believe represents the best starting point for users.
However, we emphasize that in any application, the users should compare their simulated data to the reference data set and this can be used to determine the best method for the specific application at hand, which could vary depending on data set.

@@

While we describe one special form of $\Sigma_{sim}$ where generation is highly efficient, there exist many other possible covariance matrix forms.
Block diagonal covariance matrices are one such example which can efficiently represent even high-dimensional data [@Devijver02012018] and lend themselves to efficient simulation.
Sparse matrix methods [@Bien2011-nx; @Bickel_sparse; @Cai_adaptive] yield exactly zero correlation between some variables, but it is less clear how to simulate data from these efficiently.
Iterative sampling methods [@Krylov_subspaces] are one approach that could exploit the sparse nature and may even extend to sparse inverse covariance methods [@sparse_inverse_covariance].

### 2025-06-17 manuscript/supplemental/supplemental.qmd
Edit MS and figures
And fix metabolomics figure

>>>>

```{r}
#| label: fig-compare-to-real3
#| fig-cap: "Comparison to plasma metabolomics data set, obtained from MetaboAnalyst (https://www.metaboanalyst.ca/MetaboAnalyst/upload/RocUploadView.xhtml). Methods were run using normal marginal distributions. Unlike other examples, this included more samples (59) than measured features (42 metabolites), demonstrating that these methods also work in that case. Data were log-scaled prior to simulation."
#| fig-width: 10
#| fig-height: 10
readRDS("../../processed/compare_to_real_plot_metabolomics.RDS")
```
